Parameter Recovery Validation

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: use generic parameter, posterior, and predictive sampling methods in user-facing parameter recovery validation API
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

# PyroVelocity imports
from pyrovelocity.models.modular.factory import create_piecewise_activation_model
from pyrovelocity.models.modular.data.synthetic import PiecewisePatternGenerator
from pyrovelocity.models.modular.validation import (
    PiecewiseActivationValidator,
    generate_validation_parameter_sets,
)
from pyrovelocity.plots.piecewise import (
    plot_piecewise_phase_portrait,
    plot_piecewise_trajectories,
    plot_activation_timing,
)
from pyrovelocity.plots.predictive_checks import (
    plot_prior_predictive_checks,
    plot_posterior_predictive_checks,
    plot_parameter_recovery_diagnostics,
)

@@

print(f"Created model: {model}")
print(f"Model components:")
print(f"  Dynamics: {model.dynamics_model}")
print(f"  Prior: {model.prior_model}")
print(f"  Likelihood: {model.likelihood_model}")
print(f"  Observation: {model.observation_model}")
print(f"  Guide: {model.guide_model}")
```

@@

We generate synthetic datasets representing each of the four gene expression patterns with known ground truth parameters:

```python
# Initialize pattern generator
pattern_generator = PiecewisePatternGenerator(seed=42)

# Generate validation parameter sets (2 sets per pattern = 8 total)
validation_parameter_sets = generate_validation_parameter_sets()

# Generate synthetic datasets for each parameter set
synthetic_datasets = {}
true_parameters = {}

for param_set in validation_parameter_sets:
    pattern = param_set["pattern"]
    set_id = param_set["set_id"]
    dataset_key = f"{pattern}_{set_id}"

@@

    # Generate pattern-specific data
    if pattern == "activation":
        adata = pattern_generator.generate_activation_pattern(
            num_genes=5,
            num_cells=200,
            parameters=param_set
        )
    elif pattern == "decay":
        adata = pattern_generator.generate_decay_pattern(
            num_genes=5,
            num_cells=200,
            parameters=param_set
        )
    elif pattern == "transient":
        adata = pattern_generator.generate_transient_pattern(
            num_genes=5,
            num_cells=200,
            parameters=param_set
        )
    elif pattern == "sustained":
        adata = pattern_generator.generate_sustained_pattern(
            num_genes=5,
            num_cells=200,
            parameters=param_set
        )

    # Store dataset and true parameters
    synthetic_datasets[dataset_key] = adata
    true_parameters[dataset_key] = adata.uns["true_parameters"]

    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
    print(f"  Pattern parameters: {param_set}")

print(f"\nGenerated {len(synthetic_datasets)} synthetic datasets")
```

@@

```python
# Generate prior predictive samples
print("Performing prior predictive checks...")

prior_predictive_samples = model.generate_prior_predictive_samples(
    num_samples=1000,
    num_genes=10,
    num_cells=200
)

# Plot prior predictive checks
fig_prior = plot_prior_predictive_checks(
    model=model,
    prior_samples=prior_predictive_samples,
    figsize=(15, 10)
)

@@

# Validate prior parameter ranges
print("\nPrior parameter range validation:")
for param_name, samples in prior_predictive_samples.items():
    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
        print(f"{param_name}:")
        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
```

@@

for dataset_key, adata in synthetic_datasets.items():
    print(f"\n{'='*60}")
    print(f"Validating parameter recovery for {dataset_key}")
    print(f"{'='*60}")

@@

    # Run parameter recovery validation
    result = validator.validate_parameter_recovery(
        model=model_instance,
        adata=adata,
        true_parameters=true_parameters[dataset_key],
        dataset_name=dataset_key
    )

@@

# 1. Parameter recovery diagnostic plots
fig_recovery = plot_parameter_recovery_diagnostics(
    true_parameters=true_parameters,
    inferred_parameters={k: v["posterior_samples"] for k, v in validation_results.items()},
    figsize=(20, 15)
)
plt.suptitle("Parameter Recovery Diagnostics: All Patterns", fontsize=16)
plt.savefig(viz_dir / "parameter_recovery_diagnostics.png", dpi=300, bbox_inches='tight')
plt.show()

# 2. Phase portraits for each pattern
for pattern in ["activation", "decay", "transient", "sustained"]:
    # Find datasets for this pattern
    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}

    for dataset_key, adata in pattern_datasets.items():
        fig_phase = plot_piecewise_phase_portrait(
            adata=adata,
            gene_names=adata.var_names[:3].tolist(),  # Plot first 3 genes
            posterior_samples=posterior_samples_all[dataset_key],
            true_parameters=true_parameters[dataset_key],
            figsize=(15, 10)
        )
        plt.suptitle(f"Phase Portrait: {dataset_key.replace('_', ' ').title()}", fontsize=16)
        plt.savefig(viz_dir / f"phase_portrait_{dataset_key}.png", dpi=300, bbox_inches='tight')
        plt.show()

# 3. Temporal trajectories showing piecewise dynamics
for pattern in ["activation", "transient", "sustained"]:  # Skip decay (no activation)
    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}

@@

# 4. Activation timing analysis
fig_timing = plot_activation_timing(
    posterior_samples=posterior_samples_all,
    true_parameters=true_parameters,
    figsize=(15, 10)
)
plt.suptitle("Activation Timing Analysis: All Patterns", fontsize=16)
plt.savefig(viz_dir / "activation_timing_analysis.png", dpi=300, bbox_inches='tight')
plt.show()

# 5. Posterior predictive checks for each pattern
for pattern in ["activation", "decay", "transient", "sustained"]:
    pattern_datasets = {k: v for k, v in synthetic_datasets.items() if k.startswith(pattern)}

@@

    # Generate test data with specific gamma value
    test_adata = pattern_generator.generate_activation_pattern(
        num_genes=3,
        num_cells=100,
        parameters={
            "alpha_off": 0.1,
            "alpha_on": 2.0,
            "t_on": 0.3,
            "delta": 0.4,
            "gamma_star": gamma_val
        }
    )

@@

    # Generate test data
    test_adata = pattern_generator.generate_activation_pattern(
        num_genes=config["num_genes"],
        num_cells=config["num_cells"],
        parameters=validation_parameter_sets[0]  # Use first parameter set
    )

>>>>

# PyroVelocity imports
from pyrovelocity.models.modular.factory import create_piecewise_activation_model
from pyrovelocity.models.modular.validation import (
    PiecewiseActivationValidator,
    PatternClassificationValidator,
)
from pyrovelocity.plots.piecewise import (
    plot_piecewise_phase_portrait,
    plot_piecewise_trajectories,
    plot_activation_timing,
)
from pyrovelocity.plots.predictive_checks import (
    plot_prior_predictive_checks,
    plot_posterior_predictive_checks,
    plot_parameter_recovery_diagnostics,
)

@@

print(f"Created model: {model}")
print(f"Model components:")
print(f"  Dynamics: {model.dynamics_model}")
print(f"  Prior: {model.prior_model}")
print(f"  Likelihood: {model.likelihood_model}")
print(f"  Observation: {model.observation_model}")
print(f"  Guide: {model.guide_model}")

# Demonstrate key API methods available on any PyroVelocityModel
print(f"\nKey API methods available:")
print(f"  model.sample_system_parameters() - Sample constrained or unconstrained parameters")
print(f"  model.generate_predictive_samples() - Generate prior/posterior predictive samples")
print(f"  model.train() - Train model on data")
print(f"  model.generate_posterior_samples() - Generate posterior samples after training")
print(f"  model.get_velocity() - Compute velocity estimates")
```

@@

We generate synthetic datasets representing each of the four gene expression patterns using the generic predictive sampling interface:

```python
# Generate validation datasets for all patterns using generic API
validation_datasets = {}
validation_parameters = {}

# Define pattern configurations (2 parameter sets per pattern = 8 total)
pattern_configs = [
    # Activation patterns
    {"pattern": "activation", "set_id": 1, "seed": 42},
    {"pattern": "activation", "set_id": 2, "seed": 43},
    # Decay patterns
    {"pattern": "decay", "set_id": 1, "seed": 44},
    {"pattern": "decay", "set_id": 2, "seed": 45},
    # Transient patterns
    {"pattern": "transient", "set_id": 1, "seed": 46},
    {"pattern": "transient", "set_id": 2, "seed": 47},
    # Sustained patterns
    {"pattern": "sustained", "set_id": 1, "seed": 48},
    {"pattern": "sustained", "set_id": 2, "seed": 49},
]

for config in pattern_configs:
    pattern = config["pattern"]
    set_id = config["set_id"]
    seed = config["seed"]
    dataset_key = f"{pattern}_{set_id}"

@@

    # Sample system parameters constrained to this pattern
    true_system_params = model.sample_system_parameters(
        pattern=pattern,
        set_id=set_id,
        num_samples=1,
        constrain_to_pattern=True,
        rng_key=jax.random.PRNGKey(seed)
    )

    # Generate synthetic data using the generic predictive sampling interface
    adata = model.generate_predictive_samples(
        num_cells=200,
        num_genes=5,
        samples=true_system_params,  # Use constrained parameters as "posterior"
        rng_key=jax.random.PRNGKey(seed + 100),
        return_format="anndata"
    )

    # Store dataset and true parameters
    validation_datasets[dataset_key] = adata
    validation_parameters[dataset_key] = true_system_params

    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
    print(f"  Pattern: {adata.uns['pattern_type']}")
    print(f"  True parameters shape: {[f'{k}: {v.shape}' for k, v in true_system_params.items()]}")

print(f"\nGenerated {len(validation_datasets)} synthetic datasets")
```

@@

```python
# Generate prior predictive samples using the generic interface
print("Performing prior predictive checks...")

# Generate prior predictive samples (no samples argument = sample from prior)
prior_predictive_adata = model.generate_predictive_samples(
    num_cells=200,
    num_genes=10,
    num_samples=100,  # Generate 100 prior samples
    rng_key=jax.random.PRNGKey(0),
    return_format="anndata"
)

# Extract parameter samples for validation
prior_parameter_samples = model.sample_system_parameters(
    num_samples=1000,
    constrain_to_pattern=False,  # Use full prior ranges
    rng_key=jax.random.PRNGKey(1)
)

# Plot prior predictive checks
fig_prior = plot_prior_predictive_checks(
    model=model,
    prior_adata=prior_predictive_adata,
    prior_parameters=prior_parameter_samples,
    figsize=(15, 10)
)

@@

# Validate prior parameter ranges
print("\nPrior parameter range validation:")
for param_name, samples in prior_parameter_samples.items():
    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
        print(f"{param_name}:")
        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")

# Validate that prior generates diverse expression patterns
print("\nPrior pattern diversity:")
prior_patterns = PatternClassificationValidator().classify_genes_from_parameters(
    prior_parameter_samples
)
pattern_counts = {pattern: list(prior_patterns.values()).count(pattern)
                 for pattern in ["activation", "decay", "transient", "sustained", "intermediate"]}
for pattern, count in pattern_counts.items():
    print(f"  {pattern}: {count} genes ({count/len(prior_patterns)*100:.1f}%)")
```

@@

for dataset_key, adata in validation_datasets.items():
    print(f"\n{'='*60}")
    print(f"Validating parameter recovery for {dataset_key}")
    print(f"{'='*60}")

@@

    # Run parameter recovery validation
    result = validator.validate_parameter_recovery(
        model=model_instance,
        adata=adata,
        true_parameters=validation_parameters[dataset_key],
        dataset_name=dataset_key
    )

@@

# 1. Parameter recovery diagnostic plots
fig_recovery = plot_parameter_recovery_diagnostics(
    true_parameters=validation_parameters,
    inferred_parameters={k: v["posterior_samples"] for k, v in validation_results.items()},
    figsize=(20, 15)
)
plt.suptitle("Parameter Recovery Diagnostics: All Patterns", fontsize=16)
plt.savefig(viz_dir / "parameter_recovery_diagnostics.png", dpi=300, bbox_inches='tight')
plt.show()

# 2. Phase portraits for each pattern
for pattern in ["activation", "decay", "transient", "sustained"]:
    # Find datasets for this pattern
    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}

    for dataset_key, adata in pattern_datasets.items():
        fig_phase = plot_piecewise_phase_portrait(
            adata=adata,
            gene_names=adata.var_names[:3].tolist(),  # Plot first 3 genes
            posterior_samples=posterior_samples_all[dataset_key],
            true_parameters=validation_parameters[dataset_key],
            figsize=(15, 10)
        )
        plt.suptitle(f"Phase Portrait: {dataset_key.replace('_', ' ').title()}", fontsize=16)
        plt.savefig(viz_dir / f"phase_portrait_{dataset_key}.png", dpi=300, bbox_inches='tight')
        plt.show()

# 3. Temporal trajectories showing piecewise dynamics
for pattern in ["activation", "transient", "sustained"]:  # Skip decay (no activation)
    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}

@@

# 4. Activation timing analysis
fig_timing = plot_activation_timing(
    posterior_samples=posterior_samples_all,
    true_parameters=validation_parameters,
    figsize=(15, 10)
)
plt.suptitle("Activation Timing Analysis: All Patterns", fontsize=16)
plt.savefig(viz_dir / "activation_timing_analysis.png", dpi=300, bbox_inches='tight')
plt.show()

# 5. Posterior predictive checks for each pattern
for pattern in ["activation", "decay", "transient", "sustained"]:
    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}

@@

    # Manually specify parameters with specific gamma value
    test_system_params = {
        "alpha_off": jnp.array([[0.1, 0.1, 0.1]]),
        "alpha_on": jnp.array([[2.0, 2.0, 2.0]]),
        "t_on": jnp.array([[0.3, 0.3, 0.3]]),
        "delta": jnp.array([[0.4, 0.4, 0.4]]),
        "gamma_star": jnp.array([[gamma_val, gamma_val, gamma_val]]),
        "T_max_star": jnp.array([[1.0]]),
        "t_loc": jnp.array([[0.5]]),
        "t_scale": jnp.array([[0.2]]),
    }

    # Generate test data using generic API
    test_adata = model.generate_predictive_samples(
        num_cells=100,
        num_genes=3,
        samples=test_system_params,
        rng_key=jax.random.PRNGKey(int(gamma_val * 1000)),
        return_format="anndata"
    )

@@

    # Sample parameters for benchmarking
    bench_params = model.sample_system_parameters(
        pattern="activation",
        set_id=1,
        num_samples=1,
        rng_key=jax.random.PRNGKey(42)
    )

    # Generate test data using generic API
    test_adata = model.generate_predictive_samples(
        num_genes=config["num_genes"],
        num_cells=config["num_cells"],
        samples=bench_params,
        rng_key=jax.random.PRNGKey(100),
        return_format="anndata"
    )
